Communicating causal effect heterogeneity

### 2024-06-07 ms.qmd
Manuscript revision

---
title: 'On investigating heterogeneity in psychological phenomena'
shorttitle: Heterogeneity in psychology
leftheader: Heterogeneity in psychology
author: 
  - name: Matti Vuorre
    affiliation: 1
    corresponding: yes
    address: Tilburg School of Social and Behavioral Sciences, Tilburg University
    email: mjvuorre@uvt.nl
  - name: Matthew Kay
    affiliation: 2
  - name: Niall Bolger
    affiliation: 3
affiliation:
  - id: 1
    institution: Department of Social Psychology, Tilburg School of Social and Behavioral Sciences, Tilburg University
  - id: 2
    institution: Computer Science and Communication Studies, Northwestern University
  - id: 3
    institution: Department of Psychology, Columbia University
abstract: |
  Advances in statistics and data collection methods have brought variability in psychological phenomena to the forefront of theoretical development. Yet, established methods and practices for studying heterogeneity underestimate the true uncertainty in those estimates. In the current work, we aim to deliver methods and guidelines for more rigorous treatments of heterogeneity in psychological phenomena. We provide a walkthrough of methods for i. estimating, ii. summarising, and iii. describing heterogeneity. Although other methods for incorporating estimation uncertainty are available, the methods we propose---involving bayesian hierarchical models---are especially suitable for investigating heterogeneity because they naturally describe uncertainty in all model parameters. We hope that the methods we discuss would help researchers to widen their inferential focus from statistical averages to include statistical heterogeneity in psychological phenomena.
keywords: variation, heterogeneity, hierarchical models, mixed models, methodology
authornote: |
  \textbf{This working paper is not yet peer reviewed.}
wordcount: "5,837"
bibliography: references.bib
floatsintext: yes
numbersections: false
linenumbers: no
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
documentclass: apa7
classoption: jou
header-includes:
  - |
    \usepackage{float}
    \setlength{\parskip}{0.5em}
output: 
  papaja::apa6_pdf:
    number_sections: true
    keep_tex: false
    highlight: kate
  papaja::apa6_docx: 
    number_sections: true
    keep_tex: false
---

```{r}
#| label = "setup",
#| include = FALSE
# Packages
library(papaja)
library(janitor)
library(scales)
library(ellipse)
library(knitr)
library(cmdstanr)
library(readxl)
library(lme4)
library(ggpp)
library(patchwork)
library(latex2exp)
library(ggdist)
library(ggstance)
library(tidybayes)
library(distributional)
library(posterior)
library(parameters)
library(brms)
library(tidyverse)

# Output options
opts_chunk$set(
  fig.width = 6,
  fig.align = "center",
  eval = TRUE,
  cache = TRUE,
  warning = FALSE,
  error = TRUE,
  message = FALSE
)

@@

# Model estimation options
dir.create("models", FALSE)
MAX_CORES <- as.numeric(Sys.getenv("MAX_CORES"), 4)
options(
  brms.backend = Sys.getenv("BRMS_BACKEND", "rstan"), 
  brms.threads = MAX_CORES %/% 4,
  mc.cores = MAX_CORES
)
```

When building and testing theories of psychological phenomena, psychologists have long focused on asking whether an effect exists or not. Yet, establishing that an independent variable affects a dependent variable may not be a sufficient description of the phenomenon if the treatment effect is variable from one individual to another. The importance of such between-person variation, or *heterogeneity*, for theory development is widely recognized, yet rarely communicated sufficiently in the empirical literature [@bolgerCausalProcessesPsychology2019; @brandCausalEffectHeterogeneity2013].

One reason for the scarcity of reporting and interpreting heterogeneity is that psychologists still commonly analyze data with models that obscure the assessment of heterogeneity, such as ANOVA [@bolgerCausalProcessesPsychology2019]. Although more appropriate hierarchical (or multilevel, mixed-effects) models are rapidly becoming more widespread, we believe a second important reason is that researchers do not yet have the conceptual and practical tools required to sufficiently estimate and describe heterogeneity.

When person-to-person variability is modeled and reported, those descriptions often focus on point estimates [@bolgerCausalProcessesPsychology2019], sample statistics [@beyensEffectSocialMedia2020; @vuorreThreeObjectionsNovel2022], or quantities such as the standard deviation of person-specific parameters [@bartosFairCoinsTend2023] which, in isolation, might be difficult to interpret. That is, although we are excited about the increased focus on heterogeneity, we offer that commonly used descriptions of it leave room for improvement.

Because this focus on heterogeneity in psychological phenomena is relatively new, work remains to be done in developing models, descriptions, and reporting practices that maximize investigations' impact on theory development. Our goal is to contribute to that work. We describe and illustrate the use of numerical and graphical descriptions of heterogeneity that (1) Go beyond model parameters to describe heterogeneity in clear and actionable terms, and (2) Take uncertainty in model parameters into account.

Our plan is as follows. We first review established methods for estimating and communicating expected heterogeneity of causal effects in the population using an example dataset from social psychology. We then describe additional ways in which model parameters can be interrogated and transformed to describe distributions of causal effects in the population. We then move beyond summarizing expected degrees of heterogeneity to describing distributions of plausible degrees of heterogeneity. Finally, we extend these methods to compare heterogeneity across different populations using an example dataset from cognitive psychology.

# Review of expected heterogeneity in hierarchical models

To begin our exposition, we reproduce the analyses presented in @bolgerCausalProcessesPsychology2019. In their study, which replicated findings first presented in @scholerInflatingDeflatingSelf2014, 62 participants saw twenty positively and twenty negatively valenced words, and judged whether each word was self-descriptive or not. Because most people are typically motivated to view themselves positively, @bolgerCausalProcessesPsychology2019 predicted that response times to positively valenced words would be shorter than to negatively valenced words [@scholerInflatingDeflatingSelf2014].

```{r}
#| label = "data-bolgeretal",
#| include = FALSE
dir.create("data", FALSE)
# Load, save, & clean example data from Bolger et al. (2019)
path <- "data/bolger-etal.zip"
if (!file.exists(path)) {
  download.file(
    "https://github.com/kzee/heterogeneityproject/archive/refs/heads/master.zip",
    destfile = path
  )
  unzip(path, exdir = str_remove(path, ".zip"), junkpaths = TRUE)
}

# Save variables as numerical & factor
dat <- read_csv("data/bolger-etal/heterogeneity_dataset1_traitvalence.csv") |>
  # Cleaning as in Bolger et al 2019
  filter(
    response.keys == "up",
    scale(rt) < 3,
    !(id %in% c(250, 257, 272))
  ) |>
  # Some more cleaning
  mutate(
    person = factor(id),
    trial,
    valence = factor(
      valenceE,
      levels = c(-0.5, 0.5),
      labels = c("Negative", "Positive")
    ),
    rt = logrt
  ) |>
  select(person, trial, valence, rt) |>
  arrange(person, trial)

@@

```{r}
#| label: tab-dat1
dat |>
  head() |>
  apa_table(
    span_text_columns = FALSE,
    font_size = "small",
    digits = c(0, 0, 0, 2),
    caption = "First six rows of example dataset 1 (Bolger et al., 2019; Exp 1).",
    note = "rt is log-transformed."
  )
```

In this section, we replicate @bolgerCausalProcessesPsychology2019's analysis, using their openly available data (https://github.com/kzee/heterogeneityproject). We first wrangled the data as in @bolgerCausalProcessesPsychology2019, which led to a sample of `r number(nrow(dat), big.mark = ",")` trials from 59 participants that were endorsed as self-relevant. (We present all our code in the online analysis supplement.) We show a sample of these data in Table \@ref(tab:tab-dat1). Then, we estimated the same statistical model (Model 1; equations 1.1-1.3). We modeled the log-transformed reaction time (logRT) of subject $j$ on trial $i$ as a random draw from a normal distribution with mean $\eta$ (*eta*), which could differ between trials $i$ and individuals $j$, and standard deviation $\sigma$ (*sigma*), which we assumed constant across individuals and trials, as indicated by the lack of subscripts:

\begin{equation}
\tag{1.1}
\text{logRT}_{ij} \sim \operatorname{Normal}\left(\eta_{ij}, \sigma^2\right).
\end{equation}

(We note that there are better and more informative alternatives than modeling the log-transformed RTs as normal, but those are outside the scope of this manuscript.) Then, we specified a model of the mean of the logRT distribution ($\eta_{ij}$) such that the regression coefficients captured our substantive questions:

\begin{equation}
\tag{1.2}
\eta_{ij} = \beta_0 + \gamma_{0j} + \left(\beta_1 + \gamma_{1j}\right)\text{V}_{ij}.
\end{equation}

This equation includes two kinds of parameters: $\beta_0$ (*beta*), the intercept, and $\beta_1$, the slope or effect of valence (V), do not have subscripts. In the frequentist tradition, these parameters are considered constants---not modelled on covariates---and typically referred to as "fixed" parameters [e.g. @raudenbushHierarchicalLinearModels2002]. The second set of parameters, $\gamma_{0j}$ (*gamma*) and $\gamma_{1j}$, have the subscript $j$ to indicate that they are person-specific deviations from the average intercept and slope, respectively. That is, $\beta_0 + \gamma_{01}$ is the average reaction time for person $j=1$. In frequentist texts, these are typically called "random" parameters, because they are modeled as varying randomly according to a specified distribution. Following standard multilevel modeling assumptions, we model $\gamma_0$ and $\gamma_1$ as multivariate normal distributed:

\begin{equation}
\tag{1.3}
\begin{bmatrix} 
  \gamma_0 \\ \gamma_1
\end{bmatrix} \sim 
\operatorname{MVN}\left(
  \begin{bmatrix} 0 \\ 0 \end{bmatrix}, 
  \begin{pmatrix} 
    \tau_0 & \\ 
    \rho &\tau_1 
  \end{pmatrix}
\right).
\end{equation}

In this equation, we assume that the subject-specific deviations $\gamma_0$ and $\gamma_1$ have means of zero (because the means are added to them in equation 1.2), standard deviations $\tau$ (*tau*), and a correlation $\rho$ (*rho*). Perhaps confusingly, $\tau$s and $\rho$ are also sometimes called random effects because they describe random (co)variations of the person-specific effects. To be clear, despite this naming convention they are "fixed" features of the population, not of any one group or individual.

@@

With data described in Table \@ref(tab:tab-dat1), we can estimate this model using standard maximum likelihood methods as implemented in, for example, the R package lme4 [@batesFittingLinearMixedEffects2015; @rcoreteamLanguageEnvironmentStatistical2023].

```{r}
#| label = "fit1-lmer"
fit_lmer <- lmer(
  rt ~ 1 + valence + (1 + valence | person),
  data = dat
)
```

@@

# Bootstrap CIs for variance parameters
path <- "models/lmer-fit1-bootstrap.rds"
if (!file.exists(path)) {
  boots <- confint(
    fit_lmer,
    method = "boot",
    nsim = 1000,
    oldNames = FALSE
  )
  boots <- boots[c(5, 6, 1, 3, 4, 2), ]
  tab_1[, 3:4] <- boots
  saveRDS(tab_1, path)
} else {tab_1 <- readRDS(path)}

# Get approx SEs
tab_1 <- tab_1 |> 
  mutate(SE = (CI_high - Coefficient) / 1.96) |> 
  select(Parameter, Coefficient, SE, CI_low, CI_high)
```

```{r}
#| label: tbl-fit-1
#| cache: false

m1_parnames <- c(
  "$\\beta_0$", "$\\beta_1$",
  "$\\tau_0$", "$\\tau_1$",
  "$\\rho$", "$\\sigma$"
)

tab_1 |>
  mutate(
    mutate(across(where(is.numeric), ~number(.x, .01))),
    CI = str_glue("[{CI_low}, {CI_high}]"),
    Coefficient = ifelse(
      str_starts(Coefficient, "-"), 
      Coefficient, 
      paste0("\\phantom{-}", Coefficient)
    ),
    Parameter = m1_parnames
  ) |> 
  select(
    Parameter, Coefficient, SE, CI
  ) |>
  apa_table(
    span_text_columns = FALSE,
    font_size = "small",
    caption = "Parameter estimates from Model 1.",
    escape = FALSE
  )
mu1n <- tab_1[2, 2, TRUE]
mu1n_low <- tab_1[2, "CI_low", TRUE]
sd1n <- tab_1[4, 2, TRUE]
sd1n_low <- tab_1[4, "CI_low", TRUE]
mu1 <- number(
  as.numeric(tab_1[2, c("Coefficient", "CI_low", "CI_high"), TRUE]), 
  .01
)
sd1 <- number(
  as.numeric(tab_1[4, c("Coefficient", "CI_low", "CI_high"), TRUE]), 
  .01
)
```

We show a traditional summary of the estimated parameters from this model in Table \@ref(tab:tbl-fit-1). For the average person, the estimated effect of valence on logRT is `r mu1[1]` log seconds, with a 95% bootstrap confidence interval (CI) extending from `r mu1[2]` to `r mu1[3]`. The estimated standard deviation of valence effects in the population is `r sd1[1]` log seconds, with a 95% CI of [`r sd1[2]`, `r sd1[3]`].

```{r}
#| label = "descriptors",
#| include = FALSE
# Heterogeneity interval
hi90_num <- qnorm(c(0.05, .95), mu1n, sd1n)
hi90 <- paste(
  number(hi90_num, .01),
  collapse = ", "
)
hi90_low <- paste(
  number(qnorm(c(0.05, .95), mu1n_low, sd1n_low), .01),
  collapse = ", "
)

@@

```{r}
#| label: fig-1
#| fig.height: 1.8
#| fig.env: "figure*"
#| fig.align: "center"
#| fig.cap: Heterogeneity distribution of valence effects and various descriptions of their expected heterogeneity as estimated with Model 1. A. The normal density curve defined by the point estimates of the valence effect distribution's mean ($\beta_1$) and standard deviation ($\tau_1$). Shaded areas represent areas under the normal curve within 1 (dark) and 2 (light) standard deviations of the mean. B. The 90\% Heterogeneity Interval as represented by a line segment with arrows, and the dark shaded area. C. Proportion of negative valence effects in the population (dark). D. Proportion of valence effects in the population that are within the region of practical equivalence to zero (ROPE; dark).

@@

Rows 2 and 4 in Table \@ref(tab:tbl-fit-1) define the expected normal distribution of valence effects in the population, visualized in Figure \@ref(fig:fig-1)A. In other words, our point estimate of the distribution of valence effects is Normal(`r mu1[1]`, `r sd1[1]`^2^). However, it is an incomplete description of heterogeneity for two reasons. First, it does not necessarily communicate the degree of heterogeneity in clear and actionable terms: In addition to the SD, we want quantities that describe heterogeneity in clearer terms. Below, we introduce several metrics that e.g. directly describe where a given proportion of the slopes are expected to fall. Second, it ignores the uncertainty with which the heterogeneity parameters are estimated. That is, we need ways to describe (uncertainty) distributions of distributions.

We now turn to this first issue and outline several additional metrics of heterogeneity afforded by the model's parameter estimates.

@@

First, we can use the point estimates in Table \@ref(tab:tbl-fit-1) to construct an expected *heterogeneity interval* that describes the range within which a certain percentage of the population's slopes are expected to fall [@bolgerCausalProcessesPsychology2019]. To do so, we must first determine an appropriate percentage to describe: By convention, @bolgerCausalProcessesPsychology2019 and others have focused on the 95% heterogeneity interval ($HI_{95}$). However, because there are already confusingly many quantities using the five percent cutoff, in this manuscript we focus on the 90% heterogeneity interval, and reserve 95% to describing uncertainties. The appropriate percentage to describe with a heterogeneity interval is determined by the substantive and communicative aims at hand; for our illustration 90% seemed reasonable.

To calculate a heterogeneity interval, we first specify the desired probability limits. For a 90% interval, we use .05 and .95, which together define the central 90% of the distribution. Then, we pass those probabilities and the estimated mean and standard deviation to the normal quantile function $\Phi^{-1}$ (*phi*, `qnorm()` in R), to obtain the interval: $HI_{90} = \Phi^{-1}([.05, .95], \beta_1, \tau_1) = \Phi^{-1}([.05, .95],$ `r mu1[1]`, `r sd1[1]`) = [`r hi90`]. In words, this equation calculates the 0.05 and 0.95 quantiles of the normal distribution defined by the mean's ($\beta_1$) and standard deviation's ($\tau_1$) point estimates: We expect 90% of valence effects in the population to fall within this interval. We illustrate this interval in Figure \@ref(fig:fig-1)B.

@@

In contrast, proportions of slopes above or below some critical value, or within some critical range, might serve as more informative summaries. For example, we might ask "What proportion of individuals in the population respond faster to positively valenced words?" In other words, we ask what proportion of the heterogeneity distribution is below zero. To answer, we pass zero and the estimated mean and standard deviation to the normal cumulative distribution function ($\Phi$; `pnorm()` in R): $p^- = \Phi(0, \beta_1, \tau_1) = \Phi(0,$ `r mu1[1]`, `r sd1[1]`) = `r p_less_zero`. This number---$p^-$ for proportion of population with negative effects---is the probability that a random slope from this population would take a negative value, or, in other words, the proportion of individuals in the population with negative valence effects. We illustrate this probability in Figure \@ref(fig:fig-1)C.

@@

For example, we might know from theory that valence effects in the interval [-0.1, 0.1] are practically equivalent to zero. To calculate, we can again use the normal cumulative distribution function to calculate the proportion of individuals in the population whose valence effect falls within this interval or region of practical equivalence: $p^{ROPE} = \Phi(0.1, \beta_1, \tau_1) - \Phi(-0.1, \beta_1, \tau_1)$ = `r p_rope`. In words, `r p_rope` of the population is expected to have valence effects that are practically equivalent to zero. Note that this statement's validity critically depends on the chosen interval, which we arbitrarily set as [-0.1, 0.1] in this example. We visualize this probability in Figure \@ref(fig:fig-1)D.

@@

A complementary description of heterogeneity is the proportion of the population whose effects fall above or below some critical value. For example, we can calculate proportions with negative and positive effects by using zero as the critical value. In this example, we ask "What proportion of individuals in the population have negative effects of valence?"

@@

post2 <- as_draws_df(
  fit2, 
  variable = c("b_", "sd_", "cor_"), 
  regex = TRUE
) |> 
  select(-.iteration, -.chain) |> 
  relocate(.draw, .before = b_taskFree)

>>>>

---
title: Communicating causal effect heterogeneity
shorttitle: Communicating heterogeneity
author: 
  - name: Matti Vuorre
    corresponding: true
    orcid: 0000-0001-5052-066X
    email: mjvuorre@uvt.nl
    affiliation: 
      - ref: 1
  - name: Matthew Kay
    affiliation:
      - ref: 2
  - name: Niall Bolger
    affiliation:
      - ref: 3
affiliations:
  - id: 1
    name: Tilburg University
    department: Department of Social Psychology
  - id: 2
    name: Northwestern University
    department: Computer Science and Communication Studies
  - id: 3
    name: Columbia University
    department: Department of Psychology
abstract: "Advances in experimental, data collection, and analysis methods have brought population variability in psychological phenomena to the fore. Yet, current practices for interpreting such heterogeneity underestimate the uncertainty inevitable in any statistical summary. Causal effect heterogeneity is best thought of as distribution of effects with a mean (average person’s effect) and variance (between-person differences). This average or expected heterogeneity distribution can be further summarized as a specific heterogeneity interval [@bolgerCausalProcessesPsychology2019]. However, because the underlying mean and variance parameters will be estimated with uncertainty in empirical studies, these expected distributions and intervals will underestimate the true range of effect values in the population. Using Bayesian hierarchical models, and with the aid of empirical datasets, we provide a walk-through of tools for effective heterogeneity reporting and display. We cover interval, proportion, and ratio measures of heterogeneity and their interpretation. These tools can be a spur to theory building, allowing researchers to widen their focus from population averages to population heterogeneity in psychological phenomena."
keywords: [heterogeneity, uncertainty, variation, hierarchical model, statistics]
bibliography: bibliography.bib
mainfont: Libertinus Serif
numbered-lines: true
floatsintext: true
format:
  apaquarto-pdf: 
    documentmode: doc
    keep-tex: true
    link-citations: true
  apaquarto-docx: default
---

```{r}
#| label: setup
#| include: false

# Packages
library(janitor)
library(scales)
library(ellipse)
library(cmdstanr)
library(readxl)
library(lme4)
library(ggpp)
library(patchwork)
library(latex2exp)
library(ggdist)
library(ggstance)
library(tidybayes)
library(distributional)
library(posterior)
library(parameters)
library(brms)
library(tidyverse)
library(knitr)

# Output options
opts_chunk$set(
  fig.width = 6,
  fig.align = "center",
  eval = TRUE,
  cache = TRUE,
  warning = FALSE,
  error = FALSE,
  message = FALSE
)
options(digits = 2)

@@

# Model estimation options, customize in .Renviron
dir.create("models", FALSE)
options(
  brms.backend = Sys.getenv("BRMS_BACKEND", "rstan"), 
  brms.threads = as.numeric(Sys.getenv("MAX_CORES"), 1),
  mc.cores = as.numeric(Sys.getenv("MAX_CORES"), 4)
)
```

When building and testing theories, psychologists have long focused on asking whether an effect exists or not, and what its magnitude might be. Yet, establishing that an independent variable affects a dependent variable, possibly to some specific extent, may not be a sufficient description of the phenomenon if the effect varies appreciably from one treatment unit (e.g. person) to another. The relevance of such variation in the effect, or *heterogeneity*, for theory development is recognized yet typically insufficiently described in the empirical literature [@bolgerCausalProcessesPsychology2019; @brandCausalEffectHeterogeneity2013; @gricePersonsEffectSizes2020; @richters2021].

One reason for the scarcity of reporting and sufficiently interpreting heterogeneity is that psychologists still commonly analyze data with models that obscure its assessment, such as traditional ANOVA [@bolgerCausalProcessesPsychology2019]. However, more informative modeling is not the only challenge: Although more informative hierarchical (or multilevel, mixed-effects) models are becoming widespread, most users do not yet have the conceptual and practical tools to benefit from the greater explanatory power such models afford.

When person-to-person variability is modeled and reported, those descriptions often focus on point estimates [@bolgerCausalProcessesPsychology2019], sample statistics [@beyensEffectSocialMedia2020; @vuorreThreeObjectionsNovel2022; @gricePersonsEffectSizes2020], graphical displays that don't yield numerical estimates of hypothetical data-generating mechanisms[@beck2022], or quantities such as the standard deviation of person-specific parameters [@bartosFairCoinsTend2023]. These, as we will show, provide an incomplete picture of variation: If (e.g.) a treatment effect is found for 60% or participants in a sample but the uncertainty inherent in that percentage is not communicated, we cannot make inferential conclusions about the effect's prevalence in the population. Therefore, to communicate heterogeneity effectively we need not only meaningful measures of it, but also effective methods for describing uncertainty in them. Our goal in this paper is to address this challenge by illustrating measures of heterogeneity and how to communicate it, both numerically and graphically, in ways that take uncertainty into account.

<!-- Fix quarto rendering problem in documentmode: jou. -->

<!-- \clearpage -->

<!-- Fix -->

Our plan is as follows. First, we review established methods for estimating and communicating expected heterogeneity of causal effects in the population using an example dataset from social psychology and describe additional ways in which model parameters can be transformed to describe distributions of causal effects in the population. We review the concepts and computations underlying three heterogeneity metrics: Individual effect’s mean and standard deviation; their heterogeneity interval; and the prevalence proportion. We then move beyond summarizing expected degrees of heterogeneity to describing distributions of plausible degrees of heterogeneity. Finally, we extend these methods to compare heterogeneity across different populations using an example dataset from cognitive psychology.

# Review of heterogeneity

To begin our exposition, we reproduce the analyses presented in @bolgerCausalProcessesPsychology2019. In their study, which replicated findings first presented in @scholerInflatingDeflatingSelf2014, 62 participants saw twenty positively and twenty negatively valenced words, and judged whether each word was self-descriptive or not. Because most people are typically motivated to view themselves positively, @bolgerCausalProcessesPsychology2019 predicted that responses to positively valenced words would be faster than to negatively valenced words [@scholerInflatingDeflatingSelf2014].

```{r}
#| label: data-bolgeretal
#| include: false

dir.create("data", FALSE)
# Load, save, & clean example data from Bolger et al. (2019)
path <- "data/bolger-etal.zip"
if (!file.exists(path)) {
  download.file(
    "https://github.com/kzee/heterogeneityproject/archive/refs/heads/master.zip",
    destfile = path
  )
  unzip(path, exdir = str_remove(path, ".zip"), junkpaths = TRUE)
}

# Save variables as numerical & factor
dat <- read_csv("data/bolger-etal/heterogeneity_dataset1_traitvalence.csv") |>
  # Cleaning as in Bolger et al 2019
  filter(
    response.keys == "up",
    scale(rt) < 3,
    !(id %in% c(250, 257, 272))
  ) |>
  # Some more cleaning
  mutate(
    person = fct_anon(factor(id)),
    trial,
    valence = factor(
      valenceE,
      levels = c(-0.5, 0.5),
      labels = c("Negative", "Positive")
    ),
    rt = logrt
  ) |>
  select(person, trial, valence, rt) |>
  arrange(person, trial)

@@

In this section, we replicate @bolgerCausalProcessesPsychology2019's analysis, using their openly available data. We first wrangled the data as in @bolgerCausalProcessesPsychology2019, which led to a sample of `r number(nrow(dat), big.mark = ",")` trials from `r length(unique(dat$person))` participants that were endorsed as self-descriptive. (We present our code in the online analysis supplement.) We show a sample of these data in @tbl-dat1.

```{r}
#| label: tbl-dat1
#| tbl-cap: First six rows of example dataset 1.
#| apa-note: Reaction time is log-transformed.

dat |>
  head() |>
  rename(`Reaction time` = rt) |> 
  rename_with(~str_to_sentence(.x)) |> 
  kable()
```

Then, we estimated the same statistical model (Model 1; equations 1.1-1.3). We modeled the log-transformed reaction time (logRT) of person $j$ on trial $i$ as a random draw from a normal distribution with mean $\eta$ (*eta*), which could differ between trials $i$ and individuals $j$, and standard deviation $\sigma$ (*sigma*), which we assumed constant across individuals and trials as indicated by the lack of subscripts:

```{=tex}
\begin{equation}
\tag{1.1}
\text{logRT}_{ij} \sim \operatorname{Normal}\left(\eta_{ij}, \sigma^2\right).
\end{equation}
```
(We recognize that there are better alternatives to modeling the log-transformed RTs as normal, but those are outside the scope of this manuscript.) Then, we specified a model of the mean of the logRT distribution ($\eta_{ij}$) such that the regression coefficients captured our substantive questions:

```{=tex}
\begin{equation}
\tag{1.2}
\eta_{ij} = \beta_0 + \gamma_{0j} + \left(\beta_1 + \gamma_{1j}\right)\text{V}_{ij}.
\end{equation}
```
This equation includes two sets of parameters: The first set contains $\beta_0$ (*beta*), the intercept, and $\beta_1$, the slope or effect of valence (V). Parameters in this set do not have subscripts: In the frequentist tradition, they are considered constants---not modelled on covariates---and typically referred to as "fixed" parameters [e.g. @raudenbushHierarchicalLinearModels2002]. The second set of parameters, $\gamma_{0j}$ (*gamma*) and $\gamma_{1j}$, have the subscript $j$ to indicate that they are person-specific deviations from the average intercept and slope, respectively. That is, $\beta_0 + \gamma_{01}$ is the intercept (average reaction time) for person $j=1$. In frequentist texts, these are typically called "random" parameters because they are modeled as varying randomly according to a specified distribution. Following standard multilevel modeling assumptions, we model $\gamma_0$ and $\gamma_1$ as multivariate normal distributed:

```{=tex}
\begin{equation}
\tag{1.3}
\begin{bmatrix} 
  \gamma_0 \\ \gamma_1
\end{bmatrix} \sim 
\operatorname{MVN}\left(
  \begin{bmatrix} 0 \\ 0 \end{bmatrix}, 
  \begin{pmatrix} 
    \tau_0 & \\ 
    \rho &\tau_1 
  \end{pmatrix}
\right).
\end{equation}
```
In this equation, we assume that the person-specific deviations $\gamma_0$ and $\gamma_1$ have means of zero (because the means are added to them in equation 1.2), standard deviations $\tau$ (*tau*), and a correlation $\rho$ (*rho*). Perhaps confusingly, $\tau$s and $\rho$ are also sometimes called random effects because they describe random (co)variations of the person-specific effects. To be clear, despite this naming convention they are features of the population, not of any one group or individual.

@@

With data shown in @tbl-dat1, we can estimate this model using standard (restricted) maximum likelihood methods as implemented in, for example, the R package lme4 [@batesFittingLinearMixedEffects2015; @rcoreteamLanguageEnvironmentStatistical2023].

```{r}
#| label: fit1-lmer

fit_lmer <- lmer(
  rt ~ 1 + valence + (1 + valence | person),
  data = dat
)
```

@@

# Bootstrap CIs for variance parameters
path <- "models/lmer-fit1-bootstrap.rds"
if (!file.exists(path)) {
  boots <- confint(
    fit_lmer,
    method = "boot",
    nsim = 1000,
    oldNames = FALSE
  )
  saveRDS(boots, path)
} else {
  boots <- readRDS(path)
}
tab_1[3:6, 3:4] <- boots[c(1, 3, 2, 4), ]

# Get approx SEs
tab_1 <- tab_1 |> 
  mutate(SE = (CI_high - Coefficient) / 1.96) |> 
  select(Parameter, Coefficient, SE, CI_low, CI_high)

m1_parnames <- c(
  "$\\beta_0$", "$\\beta_1$",
  "$\\tau_0$", "$\\tau_1$",
  "$\\rho$", "$\\sigma$"
)
mu1n <- tab_1[2, 2, TRUE]
mu1n_low <- tab_1[2, "CI_low", TRUE]
sd1n <- tab_1[4, 2, TRUE]
sd1n_low <- tab_1[4, "CI_low", TRUE]
mu1 <- number(
  as.numeric(tab_1[2, c("Coefficient", "CI_low", "CI_high"), TRUE]), 
  .01
)
sd1 <- number(
  as.numeric(tab_1[4, c("Coefficient", "CI_low", "CI_high"), TRUE]), 
  .01
)
```

```{r}
#| label: tbl-lmer
#| tbl-cap: Parameter estimates from Model 1 (ML).

tab_1 |>
  mutate(
    mutate(across(where(is.numeric), ~number(.x, .01))),
    `95% CI` = str_glue("[{CI_low}, {CI_high}]"),
    Parameter = m1_parnames
  ) |> 
  select(
    Parameter, Coefficient, SE, `95% CI`
  ) |>
  kable()
```

We show a conventional summary of this model's estimated parameters in @tbl-lmer. For the average person, the estimated effect of positive valence on logRT is `r mu1[1]` log seconds, with a 95% confidence interval (CI) extending from `r mu1[2]` to `r mu1[3]`. The estimated standard deviation of valence effects in the population is `r sd1[1]` log seconds. The lme4 software package does not report a standard error or CI for (co)variance parameters by default, and we therefore calculated it by bootstrapping. The resulting 95% bootstrap CI of the valence effect's standard deviation was \[`r sd1[2]`, `r sd1[3]`\].

## Heterogeneity distribution at maximum likelihood estimate of $\beta_1$ and $\tau_1$

```{r}
#| label: descriptors
#| include: false

# Heterogeneity interval
hi90_num <- qnorm(c(0.05, .95), mu1n, sd1n)
hi90 <- paste(
  number(hi90_num, .01),
  collapse = ", "
)
hi90_low <- paste(
  number(qnorm(c(0.05, .95), mu1n_low, sd1n_low), .01),
  collapse = ", "
)

@@

```{r}
#| label: fig-1
#| fig.height: 1.8
#| apa-twocolumn: true
#| fig-cap: Heterogeneity distribution of valence effects and various descriptions of their expected heterogeneity as estimated with Model 1. A. The normal density curve defined by the point estimates of the valence effect distribution's mean ($\beta_1$) and standard deviation ($\tau_1$). Shaded areas represent areas under the normal curve within 1 (dark) and 2 (light) standard deviations of the mean. B. The 90\% Heterogeneity Interval as represented by a line segment with arrows, and the dark shaded area. C. Proportion of negative valence effects in the population (dark). D. Proportion of valence effects in the population that are within the region of practical equivalence to zero (ROPE; dark).

@@

Rows 2 and 4 in @tbl-lmer define the expected normal distribution of valence effects in the population, visualized in @fig-1 A. In other words, our point estimate of the distribution of valence effects is Normal(`r mu1[1]`, `r sd1[1]`^2^). However, this distribution is an incomplete description of heterogeneity for two reasons. First, it does not incorporate uncertainty in the two determinants of heterogeneity, that is $\beta_1$ and $tau_1$: If they are precisely estimated, i.e. when uncertainty regarding them is negligible, the distribution and any quantities calculated from it characterize the population well. On the other hand, if they are estimated with considerable uncertainty, the distribution or its transformations would not characterize the population well. We return to this key issue below. Second, the distribution or its parameters do not, for many purposes, communicate heterogeneity in clear and actionable terms. Below, we introduce several metrics that e.g. directly describe where a given proportion of the slopes are expected to fall.

@@

First, we can use the point estimates in @tbl-lmer to construct an expected *heterogeneity interval* that describes the range within which a certain percentage of the population's slopes are expected to fall [@bolgerCausalProcessesPsychology2019]. To do so, we must first determine an appropriate percentage to describe: By convention, @bolgerCausalProcessesPsychology2019 and others have focused on the 95% heterogeneity interval ($HI_{95}$). However, because there are already confusingly many quantities using the five percent cutoff, in this manuscript we focus on the 90% heterogeneity interval, and reserve 95% to describing uncertainties. The appropriate percentage to describe with a heterogeneity interval is determined by the substantive and communicative aims at hand; for our illustration 90% seemed reasonable.

To calculate a heterogeneity interval, we first specify the desired probability limits. For a 90% interval, we use .05 and .95, which together define the central 90% of the distribution. Then, we pass those limits and the estimated mean and standard deviation to the normal quantile function $\Phi^{-1}$ (*phi*, `qnorm()` in R), to get the interval: $HI_{90} = \Phi^{-1}([.05, .95], \beta_1, \tau_1) = \Phi^{-1}([.05, .95],$ `r mu1[1]`, `r sd1[1]`) = \[`r hi90`\]. In words, this function calculates the 0.05 and 0.95 quantiles of the normal distribution defined by the mean's ($\beta_1$) and standard deviation's ($\tau_1$) point estimates: We expect 90% of valence effects in the population to fall within this interval. We illustrate this interval in @fig-1 B.

@@

The above HI summarizes where a given proportion of individuals' effects lie. In contrast, some applications might find it more informative to summarize proportions of effects above or below some critical value, or within some critical range. For example, we might ask "What proportion of individuals in the population respond faster to positively valenced words?" In other words, we ask a question of *prevalence*: What proportion of the heterogeneity distribution is below zero? We label this quantity $p^-$ for proportion of population with negative effects.

To answer, we pass zero (the critical value) and the estimated mean and standard deviation to the normal cumulative distribution function ($\Phi$; `pnorm()` in R): $p^- = \Phi(0, \beta_1, \tau_1) = \Phi(0,$ `r mu1[1]`, `r sd1[1]`) = `r p_less_zero`. This number is the probability that a random slope from this population would take a negative value, or, in other words, the proportion of individuals in the population with negative valence effects. We illustrate this probability in @fig-1 C.

@@

For example, let us imagine that a theory states that valence effects in the interval \[-0.1, 0.1\] are practically equivalent to zero. To calculate, we can again use the normal cumulative distribution function to calculate the proportion of individuals in the population whose valence effect falls within this interval or region of practical equivalence: $p^{ROPE} = \Phi(0.1, \beta_1, \tau_1) - \Phi(-0.1, \beta_1, \tau_1)$ = `r p_rope`. In words, `r p_rope` of the population is expected to have valence effects that are practically equivalent to zero. Note that this statement's validity critically depends on the chosen interval. We visualize this probability in @fig-1 D.

@@

A complementary description of heterogeneity is the proportion of the population whose effects fall above or below some critical value. For example, we can calculate proportions with negative and positive effects by using zero as the critical value. In this example, we ask "What proportion of individuals in the population endorse positive words faster than negative words?"

@@

post2 <- as_draws_df(
  fit2, 
  variable = c("b_", "sd_", "cor_"), 
  regex = TRUE
)
